---
title: "makeCorrLoadings() validation"
author: "Hume Winzar"
date: "April 2025"
output: rmarkdown::html_document
bibliography: references_2.bib 
link-citations: true
vignette: >
  %\VignetteIndexEntry{makeCorrLoadings() validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup-packages, include = FALSE, message=FALSE, warning=FALSE}
# Load required packages only if installed
if (requireNamespace("rosetta", quietly = TRUE)) library(rosetta)
if (requireNamespace("dplyr", quietly = TRUE)) library(dplyr)
if (requireNamespace("psych", quietly = TRUE)) library(psych)
if (requireNamespace("psychTools", quietly = TRUE)) library(psychTools)
if (requireNamespace("knitr", quietly = TRUE)) library(knitr)
if (requireNamespace("kableExtra", quietly = TRUE)) library(kableExtra)

library(LikertMakeR)
```



## Abstract

The `makeCorrLoadings()` function accurately reconstructs an item correlation 
matrix from item-factor loadings and factor correlations, similar to those 
produced in _EFA_ or _SEM_. 

The function is robust even without explicit uniqueness or factor correlation 
inputs, although loadings greater than 0.05 are needed for best results.

## The makeCorrLoadings() function

The `makeCorrLoadings()` function generates a correlation matrix from
factor loadings and factor correlations as might be published in the
results of *Exploratory Factor Analysis* (**EFA**) or a *Structural
Equation Model* (**SEM**). The resulting correlation matrix then can be
applied to the `makeItems()` function to generate a synthetic data set
of rating-scale items that closely resemble the original data that
created the factor loadings summary table.

This paper tests how well the `makeCorrLoadings()` function achieves
that goal.

## Study design

A valid `makeCorrLoadings()` function should be able to produce a
correlation matrix that is identical to the original correlation matrix.
Further, subsequent treatment with `makeItems()` should produce a
dataframe that appears to come from the same population as the original
sample.

So we need an original, _True_, dataframe to test the function. 
Preferably, we should have several different original dataframes to ensure
that results are generalisable.



## Study #1: Party Panel 2015

### Original data

We use the pp15 dataset from the `rosetta` package [@rosetta]. This is a
subset of the Party Panel 2015 dataset. Party Panel is an annual
semi-panel study among Dutch nightlife patrons, where every year, the
determinants of another nightlife-related risk behaviour is mapped. In
2015, determinants were measured of behaviours related to using highly
dosed ecstasy pills.

Nine items are relevant to this study. Each is a 7-point likert-style
question scored in the range *-3* to *+3*.

The questions and the item labels are presented as follows:

```{r}
#| echo: false

itemQuestions <- c(
  "Expectation that a high dose results in a longer trip",
  "Expectation that a high dose results in a more intense trip",
  "Expectation that a high dose makes you more intoxicated",
  "Expectation that a high dose provides more energy",
  "Expectation that a high dose produces more euphoria",
  "Expectation that a high dose yields more insight",
  "Expectation that a high dose strengthens your connection with others",
  "Expectation that a high dose facilitates making contact with others",
  "Expectation that a high dose improves sex"
)

itemLabels <- c(
  "long",
  "intensity",
  "intoxicated",
  "energy",
  "euphoria",
  "insight",
  "connection",
  "contact",
  "sex"
)

labels <- data.frame(
  Questions = itemQuestions,
  Labels = itemLabels
)

kable(labels) |>
  kable_classic(full_width = FALSE)
```

#### Extract and clean the original data file

```{r}
## variable names
item_list <- c(
  "highDose_AttBeliefs_long",
  "highDose_AttBeliefs_intensity",
  "highDose_AttBeliefs_intoxicated",
  "highDose_AttBeliefs_energy",
  "highDose_AttBeliefs_euphoria",
  "highDose_AttBeliefs_insight",
  "highDose_AttBeliefs_connection",
  "highDose_AttBeliefs_contact",
  "highDose_AttBeliefs_sex"
)

## read the data/ select desired variables/ remove obs with missing values
dat <- read.csv2(file = "data/pp15.csv") |>
  select(all_of(item_list)) |>
  na.omit()

## give variables shorter names
names(dat) <- itemLabels

sampleSize <- nrow(dat)
```

### Target correlation matrix

The correlations among these nine items should be reproducable by the
`makeCorrLoadings()` function.

```{r}
## correlation matrix
pp15_cor <- cor(dat)
```

```{r}
#| echo: false

kable(pp15_cor, digits = 2) |>
  kable_classic(full_width = FALSE)
```

#### Test cases

We shall produce the following options for factor correlation reporting:

1.  ***Full information***: factor loadings and factor correlations to
    five decimal places, plus uniquenesses.

2.  ***Full information - No uniquenesses***: factor loadings and factor
    correlations to 5 decimal places but without uniquenesses.

3.  ***Rounded loadings***: factor loadings and factor correlations to
    two decimal places, plus uniquenesses.

4.  ***Rounded loadings - No uniquenesses***: factor loadings and factor
    correlations to two decimal places but without uniquenesses.

5.  ***Censored loadings***: factor loadings to two decimal places, with
    loadings less than some arbitrary value removed for clarity of
    presentation, with uniquenesses.

6.  ***Censored loadings - No uniqueness***: factor loadings to two
    decimal places, with loadings less some arbitrary value removed for
    clarity, but without uniquenesses.

7.  ***Censored loadings, no uniqueness, no factor correlations***:
    factor loadings to two decimal places, with loadings less than some
    arbitrary value removed for clarity, no uniquenesses or factor
    correlations. Functionally, this is the equivalent of claiming
    orthogonal factors even with factor loadings from non-orthogonal
    rotation.

#### Evaluation

To compare the *True* correlation matrix with a *Synthetic* matrix, we
employ the `cortest.jennrich()` function from the `psych` package
[@psych]. This is a *Chi-square* test of whether a pair of matrices are
equal [@Jennrich1970]. We report the raw $\chi^2$ statistic and
corresponding p-value for each test.

### Exploratory Factor Analysis

Some pretesting suggest that two factors are appropriate for this
sample. And we're confident that the factors will be correlated, so we
use promax rotation.

```{r}
## factor analysis from `rosetta` package
rfaDose <- rosetta::factorAnalysis(
  data = dat,
  nfactors = 2,
  rotate = "promax"
)

factorLoadings <- rfaDose$output$loadings
factorCorrs <- rfaDose$output$correlations
```

#### Factor Loadings

```{r}
#| echo: false

kable(factorLoadings, digits = 2) |>
  kable_classic(full_width = FALSE)
```

#### Factor Correlations

```{r}
#| echo: false

kable(factorCorrs, digits = 2) |>
  kable_classic(full_width = FALSE)
```

### Test Case #1: Full information

```{r}
## round input values to 5 decimal places
# factor loadings
fl1 <- factorLoadings[, 1:2] |>
  round(5) |>
  as.matrix()
# item uniquenesses
un1 <- factorLoadings[, 3] |> round(5)
# factor correlations
fc1 <- round(factorCorrs, 5) |> as.matrix()
# run makeCorrLoadings() function
itemCors_1 <- makeCorrLoadings(
  loadings = fl1,
  factorCor = fc1,
  uniquenesses = un1
)
## Compare the two matrices
chiSq_1 <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_1,
  n1 = sampleSize, n2 = sampleSize
)
```

### Test case #2: Full information - No uniquenesses

factor loadings and factor correlations to 5 decimal places but without
uniquenesses

```{r}
## round input values to 2 decimal places
# factor loadings
fl2 <- factorLoadings[, 1:2] |>
  round(5) |>
  as.matrix()
# factor correlations
fc2 <- factorCorrs |>
  round(5) |>
  as.matrix()
itemCors_2 <- makeCorrLoadings(
  loadings = fl2,
  factorCor = fc2,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_2 <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_2,
  n1 = sampleSize, n2 = sampleSize
)
```

### Test case #3: Rounded loadings

factor loadings and factor correlations to two decimal places.

```{r}
## round input values to 2 decimal places
# factor loadings
fl3 <- factorLoadings[, 1:2] |>
  round(2) |>
  as.matrix()
# item uniquenesses
un3 <- factorLoadings[, 3] |>
  round(2)
## factor correlations
fc3 <- factorCorrs |>
  round(2) |>
  as.matrix()
## Compare the two matrices
itemCors_3 <- makeCorrLoadings(
  loadings = fl3,
  factorCor = fc3,
  uniquenesses = un3
)
## Compare the two matrices
chiSq_3 <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_3,
  n1 = sampleSize, n2 = sampleSize
)
```

### Test case #4: Rounded loadings - No uniquenesses

Factor loadings and factor correlations to two decimal places, and no
uniquenesses

```{r}
## round input values to 2 decimal places
# factor loadings
fl4 <- factorLoadings[, 1:2] |>
  round(2) |>
  as.matrix()
## factor correlations
fc4 <- factorCorrs |>
  round(2) |>
  as.matrix()
# apply the function
itemCors_4 <- makeCorrLoadings(
  loadings = fl4,
  factorCor = fc4,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_4 <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_4,
  n1 = sampleSize, n2 = sampleSize
)
```

### Censored loadings

Often, item-factor loadings are presented with lower values removed for
ease of reading. I'm calling these "Censored" loadings. This is usually
acceptable, because the purpose is to show the reader where the larger
loadings are. But such missing information may affect the results of
reverse-engineering that we're employing with the `makeCorrLoadings()`
function.

In this study, we set the level of hidden loadings at values less than
'0.1', '0.2' and '0.3'.

```{r}
#| echo: false

fl_0 <- factorLoadings[, 1:2] |>
  round(2)
fl_a <- fl_0
# convert factor loadings < '0.1' to '0'
censor_a <- 0.1
fl_a[abs(fl_a) < censor_a] <- 0
fl_a_mean <- mean(fl_a == 0, na.rm = TRUE) |> round(2)
fl_a[fl_a == 0] <- " "
fl_b <- fl_0
# convert factor loadings < '0.2' to '0'
censor_b <- 0.2
fl_b[abs(fl_b) < censor_b] <- 0
fl_b_mean <- mean(fl_b == 0, na.rm = TRUE) |> round(2)
fl_b[fl_b == 0] <- " "
fl_c <- fl_0
# convert factor loadings < '0.3' to '0'
censor_c <- 0.3
fl_c[abs(fl_c) < censor_c] <- 0
fl_c_mean <- mean(fl_c == 0, na.rm = TRUE) |> round(2)
fl_c[fl_c == 0] <- " "
#  bring factor loadings together
fl <- cbind(fl_0, fl_a, fl_b, fl_c)
colnames(fl) <- c("f1", "f2", "f1", "f2", "f1", "f2", "f1", "f2")
header_text <- c("Item" = 1, "all values" = 2, "< 0.1 out" = 2, "< 0.2 out" = 2, "<0.3 out" = 2)
# print summary factor loadings
kable(fl, digits = 2, align = rep("c", 8)) |>
  column_spec(1:9, border_left = TRUE, border_right = TRUE) |>
  kable_styling() |>
  add_header_above(
    header = header_text,
    align = "c"
  )
```

### Test case #5: Censored loadings

Factor loadings less than '0.1', '0.2', '0.3' are removed for clarity,
presented to two decimal places.

```{r}
## round input values to 2 decimal places
# factor loadings
fl5a <- factorLoadings[, 1:2] |>
  round(2)
# convert factor loadings < '0.1' to '0'
fl5a[abs(fl5a) < 0.1] <- 0
fl5a <- as.matrix(fl5a)
# item uniquenesses
un5 <- factorLoadings[, 3] |>
  round(2)
# factor correlations
fc5 <- factorCorrs |>
  round(2) |>
  as.matrix()
# apply the function
itemCors_5a <- makeCorrLoadings(
  loadings = fl5a,
  factorCor = fc5,
  uniquenesses = un5
)
## Compare the two matrices
chiSq_5a <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_5a,
  n1 = sampleSize, n2 = sampleSize
)

# factor loadings
fl5b <- factorLoadings[, 1:2] |>
  round(2)
# convert factor loadings < '0.2' to '0'
fl5b[abs(fl5b) < 0.2] <- 0
fl5b <- as.matrix(fl5b)
# apply the function
itemCors_5b <- makeCorrLoadings(
  loadings = fl5b,
  factorCor = fc5,
  uniquenesses = un5
)
## Compare the two matrices
chiSq_5b <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_5b,
  n1 = sampleSize, n2 = sampleSize
)

# factor loadings
fl5c <- factorLoadings[, 1:2] |>
  round(2)
# convert factor loadings < '0.2' to '0'
fl5c[abs(fl5c) < 0.3] <- 0
fl5c <- as.matrix(fl5c)
# apply the function
itemCors_5c <- makeCorrLoadings(
  loadings = fl5c,
  factorCor = fc5,
  uniquenesses = un5
)
## Compare the two matrices
chiSq_5c <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_5c,
  n1 = sampleSize, n2 = sampleSize
)
# kable(itemCors_5, digits = 2)
```

### Test case #6: Censored loadings - no uniquenesses

With no declared uniquenesses, they are inferred from estimated
communalities.

```{r}
itemCors_6a <- makeCorrLoadings(
  loadings = fl5a,
  factorCor = fc5,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_6a <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_6a,
  n1 = sampleSize, n2 = sampleSize
)
# apply the function
itemCors_6b <- makeCorrLoadings(
  loadings = fl5b,
  factorCor = fc5,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_6b <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_6b,
  n1 = sampleSize, n2 = sampleSize
)
# apply the function
itemCors_6c <- makeCorrLoadings(
  loadings = fl5c,
  factorCor = fc5,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_6c <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_6c,
  n1 = sampleSize, n2 = sampleSize
)
```

### Test case #7 Censored loadings, no uniqueness, no factor correlations

No uniquenesses. That is, Uniquenesses are estimated from
1-communalities, where communalities = sum(factor-loadings\^2). And no
factor correlations. That is, we assume orthogonal factors.

```{r}
itemCors_7a <- makeCorrLoadings(
  loadings = fl5a,
  factorCor = NULL,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_7a <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_7a,
  n1 = sampleSize, n2 = sampleSize
)
# apply the function
itemCors_7b <- makeCorrLoadings(
  loadings = fl5b,
  factorCor = NULL,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_7b <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_7b,
  n1 = sampleSize, n2 = sampleSize
)
# apply the function
itemCors_7c <- makeCorrLoadings(
  loadings = fl5c,
  factorCor = NULL,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_7c <- cortest.jennrich(
  R1 = pp15_cor, R2 = itemCors_7c,
  n1 = sampleSize, n2 = sampleSize
)
```

## Summary results

```{r}
#| echo: false

cases <- c(
  "Full information",
  "Full information - No uniqueness",
  "Rounded loadings",
  "Rounded loadings - No uniqueness",
  "Censored loadings <0.1",
  "Censored loadings <0.2",
  "Censored loadings <0.3",
  "Censored loadings <0.1 - no uniqueness",
  "Censored loadings <0.2 - no uniqueness",
  "Censored loadings <0.3 - no uniqueness",
  "Censored loadings <0.1 - no uniqueness, factor cors",
  "Censored loadings <0.2 - no uniqueness, factor cors",
  "Censored loadings <0.3 - no uniqueness, factor cors"
)

chi2 <- c(
  chiSq_1$chi2,
  chiSq_2$chi2,
  chiSq_3$chi2,
  chiSq_4$chi2,
  chiSq_5a$chi2,
  chiSq_5b$chi2,
  chiSq_5c$chi2,
  chiSq_6a$chi2,
  chiSq_6b$chi2,
  chiSq_6c$chi2,
  chiSq_7a$chi2,
  chiSq_7b$chi2,
  chiSq_7c$chi2
) |> round(2)

p <- c(
  chiSq_1$prob,
  chiSq_2$prob,
  chiSq_3$prob,
  chiSq_4$prob,
  chiSq_5a$prob,
  chiSq_5b$prob,
  chiSq_5c$prob,
  chiSq_6a$prob,
  chiSq_6b$prob,
  chiSq_6c$prob,
  chiSq_7a$prob,
  chiSq_7b$prob,
  chiSq_7c$prob
) |> round(3)

summary_results_1 <- data.frame(
  Treatment = cases,
  chi2 = chi2,
  p = p
)

kable(summary_results_1) |>
  kable_classic(full_width = FALSE)
```

## Conclusion

The `makeCorrLoadings` function works quite well when it has information
on factor loadings, but much less well when "summary" (censored) factor
loadings are given.

------------------------------------------------------------------------

## Study #2: Big Five (bfi personality items)

### Original data: bfi 25 Personality items representing 5 factors

25 personality self report items taken from the [International
Personality Item Pool (ipip.ori.org)](https://ipip.ori.org) were
included as part of the *Synthetic Aperture Personality Assessment*
*(SAPA)* web based personality assessment project. The data are
available through the `psychTools` package [@psychTools].

2800 subjects are included as a demonstration set for scale
construction, factor analysis, and Item Response Theory analysis. Three
additional demographic variables (sex, education, and age) are also
included.

For purposes of this study, we confine ourselves to women (sex==2) with
postgraduate qualifications (education==5), giving us a sample size of
229 subjects.

The dataset contains the following 28 variables. (The q numbers are the
SAPA item numbers).

-   A1 Am indifferent to the feelings of others. (q_146)
-   A2 Inquire about others’ well-being. (q_1162)
-   A3 Know how to comfort others. (q_1206)
-   A4 Love children. (q_1364)
-   A5 Make people feel at ease. (q_1419)
-   C1 Am exacting in my work. (q_124)
-   C2 Continue until everything is perfect. (q_530)
-   C3 Do things according to a plan. (q_619)
-   C4 Do things in a half-way manner. (q_626)
-   C5 Waste my time. (q_1949)
-   E1 Don’t talk a lot. (q_712)
-   E2 Find it difficult to approach others. (q_901)
-   E3 Know how to captivate people. (q_1205)
-   E4 Make friends easily. (q_1410)
-   E5 Take charge. (q_1768)
-   N1 Get angry easily. (q_952)
-   N2 Get irritated easily. (q_974)
-   N3 Have frequent mood swings. (q_1099
-   N4 Often feel blue. (q_1479)
-   N5 Panic easily. (q_1505)
-   O1 Am full of ideas. (q_128)
-   O2 Avoid difficult reading material.(q_316)
-   O3 Carry the conversation to a higher level. (q_492)
-   O4 Spend time reflecting on things. (q_1738)
-   O5 Will not probe deeply into a subject. (q_1964)
-   gender (Male=1, Female=2)
-   education (1=HS, 2=finished_HS, 3=some_college, 4=college_graduate
    5=graduate_degree)
-   age age in years

### Second target correlation matrix

The correlations among these 25 items should be reproducable by the
`makeCorrLoadings()` function.

```{r}
## download data
data(bfi)
## filter for highly-educated women
bfi_short <- bfi |>
  filter(education == 5 & gender == 2) |>
  na.omit()
## keep just the 25 items
bfi_short <- bfi_short[, 1:25]
sampleSize <- nrow(bfi_short)
## derive correlation matrix
bfi_cor <- cor(bfi_short)
```

#### Test cases and Evaluation

As in the first study, we shall produce the following options for factor
correlation reporting:

1.  Full information
2.  Full information - No uniquenesses
3.  Rounded loadings
4.  Rounded loadings - No uniquenesses
5.  Censored loadings
6.  Censored loadings - No uniqueness
7.  Censored loadings - no uniqueness or factor cors

And, as in the first study, we compare the *True* correlation matrix
with a *Synthetic* matrix, employing a *Jennrich test*.

### Exploratory Factor Analysis

Five correlated factors are appropriate for this sample, so we use
promax rotation.

```{r}
## factor analysis from `rosetta` package is a less messy version of the `psych::fa()` function

fa_bfi <- rosetta::factorAnalysis(
  data = bfi_short,
  nfactors = 5,
  rotate = "promax"
)

bfiLoadings <- fa_bfi$output$loadings
bfiCorrs <- fa_bfi$output$correlations
```

#### item-factor loadings & uniquenesses

```{r}
#| echo: false

kable(bfiLoadings, digits = 2) |>
  kable_classic(full_width = FALSE)
```

#### factor correlations

```{r}
#| echo: false

kable(bfiCorrs, digits = 2) |>
  kable_classic(full_width = FALSE)
```

### Test Case #1: Full information

```{r}
## round input values to 5 decimal places
# factor loadings
fl1 <- bfiLoadings[, 1:5] |>
  round(5) |>
  as.matrix()
# item uniquenesses
un1 <- bfiLoadings[, 6] |> round(5)
# factor correlations
fc1 <- round(bfiCorrs, 5) |> as.matrix()
# run makeCorrLoadings() function
itemCors_1 <- makeCorrLoadings(
  loadings = fl1,
  factorCor = fc1,
  uniquenesses = un1
)
## Compare the two matrices
chiSq_1 <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_1,
  n1 = sampleSize, n2 = sampleSize
)
```

### Other test cases as for the first study

The remaining test cases are as for the first study and this last test
case. Changes are made to the number of decimal places considered, the
presence of uniquenesses, presence of factor correlation matrix, and
level of item-factor loading that is included.

<!-- ### Test case #2: Full information - No uniquenesses  -->

<!-- factor loadings and factor correlations to 5 decimal places but without uniquenesses -->

```{r}
#| echo: false

## round input values to 5 decimal places
# factor loadings
fl2 <- bfiLoadings[, 1:5] |>
  round(5) |>
  as.matrix()
# factor correlations
fc2 <- bfiCorrs |>
  round(5) |>
  as.matrix()
# run makeCorrLoadings() function
itemCors_2 <- makeCorrLoadings(
  loadings = fl2,
  factorCor = fc2,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_2 <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_2,
  n1 = sampleSize, n2 = sampleSize
)
```

<!-- ### Test case #3: Rounded loadings -->

<!-- factor loadings and factor correlations to two decimal places. -->

```{r}
#| echo: false

## round input values to 2 decimal places
# factor loadings
fl3 <- bfiLoadings[, 1:5] |>
  round(2) |>
  as.matrix()

# item uniquenesses
un3 <- bfiLoadings[, 6] |>
  round(2)
## factor correlations
fc3 <- bfiCorrs |>
  round(2) |>
  as.matrix()
# run makeCorrLoadings() function
itemCors_3 <- makeCorrLoadings(
  loadings = fl3,
  factorCor = fc3,
  uniquenesses = un3
)

## Compare the two matrices
chiSq_3 <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_3,
  n1 = sampleSize, n2 = sampleSize
)
```

<!-- ### Test case #4: Rounded loadings - No uniquenesses -->

<!-- Factor loadings and factor correlations to two decimal places, and no uniquenesses -->

```{r}
#| echo: false

## round input values to 2 decimal places
# factor loadings
fl4 <- bfiLoadings[, 1:5] |>
  round(2) |>
  as.matrix()
## factor correlations
fc4 <- bfiCorrs |>
  round(2) |>
  as.matrix()
# run makeCorrLoadings() function
itemCors_4 <- makeCorrLoadings(
  loadings = fl4,
  factorCor = fc4,
  uniquenesses = NULL
)

## Compare the two matrices
chiSq_4 <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_4,
  n1 = sampleSize, n2 = sampleSize
)
```

<!-- ### Test case #5: Censored loadings -->

<!-- Factor loadings less than '0.3' are removed for clarity, presented to two decimal places. Uniquenesses are inferred from estimated communalities. -->

```{r}
#| echo: false

## round input values to 2 decimal places
# factor loadings
fl5a <- bfiLoadings[, 1:5] |>
  round(2)
# convert factor loadings < '0.1' to '0'
fl5a[abs(fl5a) < 0.1] <- 0
fl5a <- as.matrix(fl5a)
# factor correlations
fc5 <- bfiCorrs |>
  round(2) |>
  as.matrix()
# item uniquenesses
un5 <- bfiLoadings[, 6] |>
  round(2)
# run makeCorrLoadings() function
itemCors_5a <- makeCorrLoadings(
  loadings = fl5a,
  factorCor = fc5,
  uniquenesses = un5
)
## Compare the two matrices
chiSq_5a <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_5a,
  n1 = sampleSize, n2 = sampleSize
)
# factor loadings
fl5b <- bfiLoadings[, 1:5] |>
  round(2)
# convert factor loadings < '0.2' to '0'
fl5b[abs(fl5b) < 0.2] <- 0
fl5b <- as.matrix(fl5b)
# run makeCorrLoadings() function
itemCors_5b <- makeCorrLoadings(
  loadings = fl5b,
  factorCor = fc5,
  uniquenesses = un5
)
## Compare the two matrices
chiSq_5b <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_5b,
  n1 = sampleSize, n2 = sampleSize
)
# factor loadings
fl5c <- bfiLoadings[, 1:5] |>
  round(2)
# convert factor loadings < '0.3' to '0'
fl5c[abs(fl5c) < 0.3] <- 0
fl5c <- as.matrix(fl5c)
# run makeCorrLoadings() function
itemCors_5c <- makeCorrLoadings(
  loadings = fl5c,
  factorCor = fc5,
  uniquenesses = un5
)
## Compare the two matrices
chiSq_5c <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_5c,
  n1 = sampleSize, n2 = sampleSize
)
```

<!-- ### Test case #6: Censored loadings - No uniquenesses -->

<!-- Factor loadings less than '0.1', '0.2', '0.3' are removed for clarity, presented to two decimal places. -->

```{r}
#| echo: false

## Loadings and factor correlations are the same, so we only need to change
## parameters of the makeCorrLoadings() application.

# run makeCorrLoadings() function
itemCors_6a <- makeCorrLoadings(
  loadings = fl5a,
  factorCor = fc5,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_6a <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_6a,
  n1 = sampleSize, n2 = sampleSize
)
# run makeCorrLoadings() function
itemCors_6b <- makeCorrLoadings(
  loadings = fl5b,
  factorCor = fc5,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_6b <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_6b,
  n1 = sampleSize, n2 = sampleSize
)
# run makeCorrLoadings() function
itemCors_6c <- makeCorrLoadings(
  loadings = fl5c,
  factorCor = fc5,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_6c <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_6c,
  n1 = sampleSize, n2 = sampleSize
)
```

<!-- ### Test case #7 Censored loadings, no uniqueness, no factor correlations -->

<!-- Factor loadings less than '0.2' are removed for clarity, but no uniquenesses. That is, Uniquenesses are estimated from 1-communalities, where communalities = sum(factor-loadings\^2). -->

```{r}
#| echo: false

## Loadings and factor correlations are the same, so we only need to change
## parameters of the makeCorrLoadings() application.

# run makeCorrLoadings() function
itemCors_7a <- makeCorrLoadings(
  loadings = fl5a,
  factorCor = NULL,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_7a <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_7a,
  n1 = sampleSize, n2 = sampleSize
)
# run makeCorrLoadings() function
itemCors_7b <- makeCorrLoadings(
  loadings = fl5b,
  factorCor = NULL,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_7b <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_7b,
  n1 = sampleSize, n2 = sampleSize
)
# run makeCorrLoadings() function
itemCors_7c <- makeCorrLoadings(
  loadings = fl5c,
  factorCor = NULL,
  uniquenesses = NULL
)
## Compare the two matrices
chiSq_7c <- cortest.jennrich(
  R1 = bfi_cor, R2 = itemCors_7c,
  n1 = sampleSize, n2 = sampleSize
)
```

## Study #2 Summary results

```{r}
#| echo: false

cases <- c(
  "Full information",
  "Full information - No uniquenesses",
  "Rounded loadings",
  "Rounded loadings - No uniquenesses",
  "Censored loadings <0.1",
  "Censored loadings <0.2",
  "Censored loadings <0.3",
  "Censored loadings <0.1 - no uniqueness",
  "Censored loadings <0.2 - no uniqueness",
  "Censored loadings <0.3 - no uniqueness",
  "Censored loadings <0.1 - no uniqueness, no factor cors",
  "Censored loadings <0.2 - no uniqueness, no factor cors",
  "Censored loadings <0.3 - no uniqueness, no factor cors"
)

chi2 <- c(
  chiSq_1$chi2,
  chiSq_2$chi2,
  chiSq_3$chi2,
  chiSq_4$chi2,
  chiSq_5a$chi2,
  chiSq_5b$chi2,
  chiSq_5c$chi2,
  chiSq_6a$chi2,
  chiSq_6b$chi2,
  chiSq_6c$chi2,
  chiSq_7a$chi2,
  chiSq_7b$chi2,
  chiSq_7c$chi2
) |> round(2)

p <- c(
  chiSq_1$prob,
  chiSq_2$prob,
  chiSq_3$prob,
  chiSq_4$prob,
  chiSq_5a$prob,
  chiSq_5b$prob,
  chiSq_5c$prob,
  chiSq_6a$prob,
  chiSq_6b$prob,
  chiSq_6c$prob,
  chiSq_7a$prob,
  chiSq_7b$prob,
  chiSq_7c$prob
) |> round(3)

summary_results_2 <- data.frame(
  Treatment = cases,
  chi2 = chi2,
  p = p
)

# summary_results_2
kable(summary_results_2, digits = c(0, 1, 5)) |>
  kable_classic(full_width = FALSE)
```

# Overall Results

Both studies show consistent results.

```{r}
#| echo: false

overall_summary <- data.frame(
  treatment = cases,
  chi2.1 = summary_results_1[, 2],
  p.1 = summary_results_1[, 3],
  chi2.2 = summary_results_2[, 2],
  p.2 = summary_results_2[, 3]
)

names(overall_summary) <- c("treatment", "chi2", "p", "chi2", "p")

kable(overall_summary, digits = c(0, 1, 3, 1, 3)) |>
  column_spec(4, border_left = TRUE) |>
  kable_classic(full_width = FALSE) |>
  add_header_above(c(" " = 1, "Party panel" = 2, "Big 5 (bfi)" = 2))
```

## Conclusions

The `makeCorrLoadings()` function is designed to produce an item
correlation matrix based on item-factor loadings and factor correlations
as one might see in the results of Exploratory Factor Analysis (EFA) or
Structural Equation Modelling (SEM).

Results of this study suggest that the `makeCorrLoadings()` function
does a surprisingly good job of reproducing the target correlation
matrix when all item-factor loadings are present.

A correlation matrix created with `makeCorrLoadings()` seems to be
robust even with the absence of specified *uniquenesses*, or even
without *factor correlations*. But a valid reproduction of a correlation
matrix should have complete item-factor loadings.

------------------------------------------------------------------------

## References
