---
title: "makeItemsScale() explainer"
author:
  - name: Hume Winzar
    id: hw
    orcid: 0000-0001-7475-2641
    email: hume.winzar@mq.edu.au
    affiliation: 
      - name: Macquarie University
        city: Sydney
        state: NSW
        country: Australia
        url: www.mq.edu.au

abstract: | 
   The makeItemsScale() function reconstructs Likert item responses from scale scores.  It selects candidate item partitions whose dispersion matches the correlation implied by a target Cronbach’s alpha. The resulting dataset reproduces the scale scores while closely matching the requested reliability.

format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: TRUE
    toc-depth: 2  
  pdf:
    documentclass: scrartcl
    classoption: DIV=11
    papersize: A4
  
---

# Reconstructing Likert Items from Scale Scores with a Target Reliability

In many situations researchers have access to scale scores but not the individual item responses that produced them. For example, published studies often report only summary statistics, simulation studies may require synthetic item data, and teaching examples may benefit from realistic item responses. Generating such data is not straightforward because the reconstructed items must satisfy two constraints simultaneously: the item values must sum to the observed scale score, and the items should exhibit a desired level of reliability.

The `makeItemsScale()` function addresses this problem by reconstructing plausible Likert-style item responses whose sums reproduce the supplied scale scores while approximating a target _Cronbach’s alpha_. The algorithm achieves this by selecting candidate item combinations whose dispersion corresponds to the inter-item correlation implied by the desired reliability.

Reconstructed item-level Likert responses must satisfy two constraints:

 1. The item values must sum to a given scale score.
 2. The items should exhibit a desired reliability (Cronbach’s $\alpha$).
 
```{r}
#| label: "load_libraries"
#| warning: false
#| message: false
#| echo: TRUE

## load required packages
library(dplyr)
library(ggplot2)
library(ggrepel)
library(gtools)
library(knitr)
library(kableExtra)
library(LikertMakeR)
```


Quick example:

The short dataset in @tbl-opening_example shows a four-item 5-point Likert 
scale, `myScale` for which we want to produce scale items that match the 
scores in the dataset and have _Cronbach's alpha_ of `0.80`. 
This result is achieved in the four columns, `V1 ... V4` which average to 
equal the corresponding values in `myScale`.


<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: tbl-opening_example
#| tbl-cap: "Short Example: 4-item 5-point Likert scale, alpha = 0.8"
#| warning: false
#| message: false
#| echo: TRUE
n <- 16
mean <- 3
sd <- 1.0
lower <- 1
upper <- 5
k <- 4

target_alpha <- 0.8

df <- data.frame(
  myScale = lfast(
    n = n, mean = mean, sd = sd,
    lowerbound = lower, upperbound = upper,
    items = k
  )
)

myItems <- makeItemsScale(
  scale = df,
  lowerbound = lower,
  upperbound = upper,
  items = k,
  alpha = target_alpha,
  summated = FALSE
)

myAlpha <- alpha(, myItems) |> round(4)

df <- cbind(df, myItems)

knitr::kable(df)

```

</div>

Here, the resulting _Cronbach's alpha_ = `r myAlpha`, so the synthetic data 
are correct to two decimal places. Not bad for just 16 observations! 
_(Actually, number of observations has little to do with **alpha**)_

The `makeItemsScale()` function generates such data by selecting candidate 
item combinations whose dispersion matches the correlation structure implied 
by the target reliability.

The key idea is that rows with similar item values produce stronger 
correlations between items, while rows with widely varying values produce 
weaker correlations. 

## Relationship between reliability and correlation

Cronbach’s alpha depends on the **average inter-item correlation**:


$$
\alpha = \frac{k \bar r}{1 + (k - 1) \bar r}
$$

where

  - $k$ = number of items
  - $\bar r$  = mean inter-item correlation.

Rearranging gives the correlation implied by a desired reliability:


$$
\bar r = \frac{\alpha}{k - \alpha (k - 1)}
$$


### Example 

Suppose we want

  - 4 items
  - target $\alpha$ = 0.80

Then

$$
\bar r = \frac{0.80}{4 - 0.80 (3)} = 0.5
$$

Typically, we would formalise this calculation with a function.

```{r}
#| label: "alpha_2_r_function"
#| echo: TRUE

alpha_2_r <- function(target_alpha, k) {
  mean_r <- target_alpha / (k - target_alpha * (k - 1))
  
  return(mean_r)
}

```


So the reconstructed items should exhibit an average correlation of 
approximately $0.50$.


## Candidate rows

The algorithm first generates all possible item combinations within the 
response bounds.

For a 4-item 1–5 scale, there are 70 possible rows (aka: _combinations with replacement_), as shown in @tbl-head_candidates.

```{r}
#| label: "candidates"
#| echo: TRUE
#| 
lower <- 1
upper <- 5
k <- 4


candidates <- gtools::combinations(
  v = c(lower:upper),
  r = k,
  n = length(c(lower:upper)),
  repeats.allowed = TRUE
) |>
  data.frame()

```




<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: tbl-head_candidates
#| tbl-cap: "Candidate rows"
#| echo: false

sep <- candidates[1, ]
sep[,] <- "..."

out <- rbind(
  head(round(candidates, 4)),
  sep,
  tail(round(candidates, 4))
)

knitr::kable(out)
```

</div>


A scale made from six 5-point items has 210 possible combinations. 
A scale made from three 7-point items has 84 possible combinations.


Each candidate row represents a possible pattern of item responses.

Each row sums to a value that could be appear in a summated scale. 
Similarly, each row shows variation in its values, as measured by the 
row standard deviation

$$
s = \sqrt{\frac{1}{k - 1}\sum^k_{i=1}  (x_i - \bar x)^2}
$$

Rows with small $s$ contain similar item values, implying stronger correlations.

Rows with large $s$ contain more varied values, implying weaker correlations.

```{r}
#| label: "candidate_moments"
#| echo: TRUE

candidates$sum <- rowSums(candidates[, 1:k])
candidates$sd <- apply(X = candidates[, 1:k], MARGIN = 1, FUN = sd)

```


<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: tbl-head_candidate_moments
#| tbl-cap: "sum and sd of candidate rows"
#| echo: false

sep <- candidates[1, ]
sep[,] <- "..."

out <- rbind(
  head(round(candidates, 4)),
  sep,
  tail(round(candidates, 4))
)

knitr::kable(out)
```

</div>

### Similarity index

To relate row dispersion to correlation, the algorithm converts the standard 
deviation into a **similarity index**

$$
r^* = 1 - \frac{s}{s_{max}}
$$

where

  - $s$ = row standard deviation
  - $s_{max}$ = maximum dispersion observed among candidate rows.

This transformation maps dispersion onto a scale between  $0$ (low similarity) 
and $1$ (high similarity).

The similarity index is calculated using the maximum dispersion observed 
across all possible item combinations within the response range. 

Rows with similar values therefore have large $r^*$

```{r}
#| label: "candidate_similarities"
#| echo: TRUE

s_max <- max(candidates$sd)

candidates$similar <-  1 - candidates$sd / s_max

```


<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: tbl-show_candidate_similarities
#| tbl-cap: "Top and bottom rows of candidate similarities"
#| echo: false

sep <- candidates[1, ]
sep[,] <- "..."

out <- rbind(
  head(round(candidates, 4)),
  sep,
  tail(round(candidates, 4))
)

knitr::kable(out)
```

</div>

### Selecting candidate rows

For each candidate row we compute the absolute difference between the target 
mean correlation coefficient and the row similarity score.

$$
| {r^* - \bar r} |
$$

where

  - $r^*$ = row similarity
  - $\bar r$ = target mean correlation.

Rows whose similarity index is **closest to the target correlation** are 
preferred.

## Visual explanation of candidate selection

@fig-similarity_sum_graphic shows all 70 possible combinations of 
four 5-point items. Each point represents one combination of four items, 
arranged by the sum of items and the similarity score. 
Sum-of-items ranges from '4' `1111` to '20' `5555`, 
and the similarity score ranges from '0' 
(minimum similarity/ maximum variance - here representing `1155` to '1' 
(perfect similarity/ zero variance). 
Note the five points at the top, at similarity = '1' that represent the 
combinations where all item values are equal. 

We find the "best" combination at the intersection of the given sum and our 
target alpha. Recall that, with four items, a target alpha of '0.80' 
corresponds to a mean correlation of '0.5'. 

For example, if we have a scale that sums to '8', we see there are five 
combinations with that sum: `2222`, `1223`, `1133`, `1124`, and `1115`, with 
progressively decreasing similarity score. With a target alpha of '0.80' then 
we select combination `1133`, (similarity='0.5'). If target alpha is '0.70', 
corresponding to similarity='0.368' the closest combination is `1124`, 
so that is chosen. And we choose `1223` if target alpha is '0.90' 
(similarity='0.692'). 



```{r}
#| label: fig-similarity_sum_graphic
#| fig-cap: "Possible combinations of four 5-point items. Select best combination at the intersection of given sum and desired alpha."
#| fig-width: 7
#| fig-height: 6
#| out-width: "100%"
#| echo: FALSE

candidates$label <- apply(
  candidates[, c("X1","X2","X3","X4")],
  1,
  function(x) paste0("(", paste(x, collapse = ""), ")")
)


ggplot(candidates, aes(x = sum, y = similar)) +
  geom_jitter(
    size = 3,
    colour = "#0077B3", alpha = 0.3,
    width = 0.0, height = 0.0
  ) +
  geom_text_repel(aes(label = label), size = 2, alpha = 1) +
  scale_x_continuous(name = "Sum of scale items", breaks = c(4,8,12,16,20)) +
  scale_y_continuous(name = "Similarity score (r*)") +
  geom_hline(yintercept = 0.5, colour = "#FF6F61") +
  annotate("text", x = min(candidates$sum) - 0.5, y=0.52,
           label = "alpha == 0.80",
           parse = TRUE, size=3) +
  geom_hline(yintercept = 0.69, colour = "#FFB300") +
  annotate("text", x = min(candidates$sum) - 0.5, y=0.71,
           label = "alpha == 0.90",
           parse = TRUE, size=3) +
  geom_hline(yintercept = 0.37, colour = "#FFB300") +
  annotate("text", x = min(candidates$sum) - 0.5, y=0.39,
           label = "alpha == 0.70",
           parse = TRUE, size=3) +
  theme_minimal() + 
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )


```



## Worked example 1

**Four 5-point items with summed score = 12**

Suppose a respondent’s scale score is **12**.

We must find item values such that

$$
x_1 + x_2 + x_3 +x_4 = 12
$$

Possible combinations include:

<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: "sums_12"
#| echo: TRUE

sums_12 <- filter(candidates, sum == 12)

print(sums_12)

```

</div>


Recall the target correlation is **0.50**.

The row

`2  2  4  4`

has similarity **0.50**, which is closest to the target similarity score.
So this row is selected (and later permuted across item positions).
You can verify this in @fig-similarity_sum_graphic.


## Worked example 2

**Four 5-point items with summed score = 7**

Now consider a scale score of **7**.

Possible rows include:

<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: "sums_7"
#| echo: TRUE

sums_7 <- filter(candidates, sum == 7)

print(sums_7)

```

</div>

The row

`1  1  2  3  ` 

has similarity closest to the target similarity (0.50), so it is selected.
Check also in @fig-similarity_sum_graphic.


## Constructing the dataset

The algorithm repeats this process for every scale score.

  1. Identify candidate rows that sum to the required score
  2. Compute dispersion for each candidate
  3. Convert dispersion to similarity index
  4. Select the row whose similarity best matches the target correlation
  5. Randomly permute item positions

Finally, a short optimisation step rearranges item values within rows to 
improve the overall correlation structure while preserving each row sum.

## The final optimisation step

The short optimisation step takes the selected rows that correspond to the 
desired scale sums and iteratively rearranges values within each row until 
the desired Cronbach's alpha is achieved. 

This step implements a stochastic local search procedure in which small, 
random perturbations are applied to individual rows and retained only if they 
improve the fit to the target reliability.

The algorithm follows the following steps:

 1. Initialise the current dataset as the best dataset.
 2. Repeat:
    - Randomly select a row.
    - Randomly select two positions within that row and swap their values.
    - Recalculate Cronbach’s alpha for the updated dataset.
    - If the new alpha is closer to the target value, retain the change (i.e., update the best dataset); otherwise, revert the swap.
 3. Stop when the alpha falls within a specified tolerance of the target; otherwise, continue iterating.


The effect of the optimisation step is best shown with an example. 
Consider a scale similar to that shown at the top of this 
explainer, @tbl-selectItems, this time with eight 
observations for brevity.

```{r}
#| label: optimisation_setup
#| message: false
#| echo: FALSE

# set.seed(123)

n <- 8
mean <- 3
sd <- 1.0
lower <- 1
upper <- 5
k <- 4

target_alpha <- 0.8

df <- data.frame(
  myScale = LikertMakeR::lfast(
    n = n, mean = mean, sd = sd,
    lowerbound = lower, upperbound = upper,
    items = k
  )
)

trueItems <- LikertMakeR::makeItemsScale(
  scale = df,
  lowerbound = lower,
  upperbound = upper,
  items = k,
  alpha = target_alpha,
  summated = FALSE
)

alpha_optim <- LikertMakeR::alpha(, trueItems) |> round(3)

randItems <- trueItems
for (i in seq_len(nrow(randItems))) {
  randItems[i, ] <- sample(as.numeric(randItems[i, ]))
}

selectItems <- trueItems
for (i in seq_len(nrow(selectItems))) {
  selectItems[i, ] <- sort(as.numeric(selectItems[i, ]))
}

myscale <- data.frame(
  scale = rowMeans(selectItems),
                      sums = rowSums(selectItems)
                      )

selectItems <- cbind(myscale, selectItems)

alpha_random <- LikertMakeR::alpha(, randItems) |> round(3)

cor_random <- cor(randItems) |> round(2)
cor_optim <- cor(trueItems) |> round(2)

mean_r_optim <- alpha_optim / (k - alpha_optim * (k - 1))

mean_r_random <- alpha_random / (k - alpha_random * (k - 1))

```


The function `LikertMakeR::makeItemsScale()` accepts a given vector of 
Likert-scale values, as either means or as a summated scale, and then 
extracts appropriate item combinations for the desired Cronbach's alpha.


<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: tbl-selectItems
#| tbl-cap: "Given Scale (mean=3, sd=1), and selected item combinations for target alpha=0.8"
#| warning: false
#| message: false
#| echo: FALSE

kable(selectItems)

```

</div>

As part of the same process, as shown in @tbl-randItems, the values within each selected item combination are randomly rearranged.


<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: tbl-randItems
#| warning: false
#| message: false
#| echo: FALSE

cap_rand <- paste0(
  "Derived scale items before optimisation ($\\alpha$ = ", 
  alpha_random, 
  ")"
)

kable(randItems, caption = cap_rand)

```

</div>


Comparison of the initial scale reconstruction (@tbl-selectItems), and then the randomised scales (@tbl-randItems) and optimised scales (@tbl-trueItems) show identical row values, but only the optimised arrangement achieves the target reliability.


<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: tbl-trueItems
#| warning: false
#| message: false
#| echo: FALSE

cap_optim <- paste0(
  "Derived scale items after optimisation ($\\alpha$ = ", 
  alpha_optim, 
  ")"
)

kable(trueItems, caption = cap_optim)

```

</div>


Correlation matrices of the data, before and after the alpha-search 
optimisation step, are presented in @tbl-cor_random and @tbl-cor_optim.

We see that randomly-allocated row values produce mean correlations well-below 
that required to achieve the desired alpha, but after optimisation the values 
are typically correct within two decimal places.


<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: tbl-cor_random
#| echo: false

cap_cor_random <- paste0(
  "Item correlations before optimisation ($\\alpha$ = ", 
  alpha_random, 
  ")"
)

kable(
  cor_random, 
  caption = cap_cor_random,
  align = "c") |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )

```

</div>



<div style="width: fit-content; margin-left: auto; margin-right: auto;">

```{r}
#| label: tbl-cor_optim
#| echo: false

cap_cor_optim <- paste0(
  "Item correlations after optimisation ($\\alpha$ = ", 
  alpha_optim, 
  ")"
)

kable(
  cor_optim, 
  caption = cap_cor_optim,
  align = "c") |>
  kable_styling(
    full_width = FALSE
  )
```

</div>


## Why this works

Cronbach’s alpha depends on the average correlation between items across respondents. By selecting candidate rows whose within-row dispersion corresponds to the desired correlation level, the algorithm indirectly controls the covariance structure of the generated dataset.

Rows with similar item values generate stronger correlations, while rows with more varied values generate weaker correlations. By selecting rows with the appropriate similarity level for each scale score, the algorithm produces datasets whose observed reliability closely matches the requested value.

Many different candidate rows share identical dispersion values. When this occurs at the same sum-value, the algorithm randomly selecting among tied candidates. This avoids systematic preference for particular partitions while preserving the desired correlation structure.

Candidate item partitions produce a discrete set of dispersion values, which means the achieved reliability can only approximate the requested value. In simulation tests the deviation from the target Cronbach’s $\alpha$ is typically less than `0.001`.


